D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem

### 2024-01-10 ootavacuous.tex
Add "conjecture" weasel words  ;-)
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by C++
	implementers, and this paper summarizes some of the challenges.
	But at the same time, there are no known instances of OOTA
	behavior in real C++ implementations.

	This paper refactors the commonly used condition for detecting
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and on hardware systems.
	These constraints prevent OOTA cycles containing only volatile
	atomics from occurring in UB-free executions of C++ programs
	running on correctly constructed C++ single-thread analysis
	implementations in our current universe.
	Given current optimizations, these constraints also prevent
	OOTA cycles involving non-volatile atomics.
\end{abstract}

@@

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This reasoning shows that correct real-world C++ implementations relying
on single-threaded analyses are inherently incapable of producing OOTA
cycles in UB-free executions of programs using volatile atomics.

C++ implementations that additionally refrain from inventing atomic loads
and that also ensure that each atomic load that is executed corresponds
to a specific load in the abstract machine are incapable of producing
OOTA cycles in UB-free executions of programs using non-volatile as well
as volatile atomics.

@@

Please note that the OOTA-like cycle called out in the \co{exists}
clause can happen even though the program never actually executes any
code that would have resulted in division-by-zero UB.
It is tempting to consider using the techniques described in
Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the C++ implementation can detect and exploit UB when
considering line~14 in isolation.
Once the constant 1 is substituted for line~14's expression, the
hardware is free to reorder lines~13 and~14 even if a compiler-based
C++ implementation refrains from doing so.

>>>>

\begin{abstract}
	The out-of-thin-air (OOTA) properties of the specification
	of \co{memory_order_relaxed} have resulted in considerable
	consternation over the years.
	Attempts to create memory models that rule out OOTA behaviors
	have either been non-executable, complex, or unloved by C++
	implementers, and this paper summarizes some of the challenges.
	But at the same time, we know of no instances of OOTA behavior
	in real C++ implementations.

	This paper refactors the commonly used condition for classifying
	OOTA behavior, which permits bringing to bear constraints from
	long-standing laws of physics and from hardware systems.
	We present informal arguments that these constraints prevent
	OOTA cycles containing only volatile atomics from occurring
	in UB-free C++ programs running on correctly constructed C++
	implementations that are restricted to single-threaded analysis
	in our current universe.
	We further conjecture that, given current optimizations, these
	constraints also prevent OOTA cycles involving non-volatile
	atomics.
\end{abstract}

@@

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This reasoning shows that correct real-world C++ implementations relying
on single-threaded analyses are inherently incapable of producing OOTA
cycles in UB-free programs using volatile atomics.

C++ implementations that additionally refrain from inventing atomic loads
and that also ensure that each atomic load that is executed corresponds
to a specific load in the abstract machine are incapable of producing
OOTA cycles in UB-free programs using non-volatile as well
as volatile atomics.

@@

Please note that the OOTA-like cycle called out in the \co{exists}
clause can happen even though the program never actually executes any
code that would have resulted in division-by-zero UB.
It is tempting to consider using the techniques described in
Appendix~\ref{app:Aside on Undefined Behavior}
that prevent back-propagation of UB, but these are ineffective in this
case because the C++ implementation can detect and exploit UB when
considering line~14 in isolation.
Once the constant 1 is substituted for line~14's expression, the
hardware is free to reorder lines~13 and~14 even if a compiler-based
C++ implementation refrains from doing so.
Note that this OOTA-like cycle can appear even if all of the atomic
operations in
Listing~\ref{lst:OOTA Due To Divide-By-Zero UB}
were converted to be volatile.
